Mendelian randomization with genome-wide association studies data

### 2022-04-27 reply-to-review1/mr-review.Rmd
added citations for a few more sections

## Methods with Summary data
The proliferation of publicly available GWAS data accelerates the development of MR methods with multiple SNP instruments that make use of GWAS summary statistics for model fitting. Our timeline highlights some of the many multi-instrument MR methods with GWAS summary statistics, with an increasing density of methods over time in the last three years (Figure 2). These methods include the inverse variance weighted MR [61], MR-Egger [51], weighted median estimation [68], Bayesian weighted MR [69], robust adjusted profile score [70], MRMix [71], CAUSE [72], and MRAID [73]. These methods differ in their approaches to three considerations that include instrument selection (from among the available SNPs), modeling and controlling for horizontal pleiotropy, and statistical inference procedure. We summarize some of these MR methods that use independent or correlated SNP instruments below. We will cover the remaining MR methods that model horizontal pleiotropy in a later section.

Figure 2: Two-sample MR with GWAS summary statistics methods publications density increases over time.
[61] termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
Specifically, [61] combined causal effect ratio estimates from  independent SNPs by using Equation (5.1).

[68], noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
weighted median-based estimators, building on research from [94]. Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [95]. Specifically, let  denote the  ordered ratio estimate statistic, from least to greatest. If an odd number, say , of SNP instruments is used, then the simple median estimator is defined as the  ordered ratio estimate. For an even number, say , number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, .
Due to the inefficiency of the simple median estimator, [68] define the weighted median estimator. To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator. The weighted median estimator, then is the median of a distribution defined by having  as the  percentile, where  is the weight assigned to the  ordered ratio estimate. In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
[68] also studied the penalized weighted median estimator. They defined the penalized weights, , where  is the p-value resulting from the comparison of  to a  distribution [68,96]. The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [14,61–63]. In addition, LD clumping for choosing instrument SNPs can be problematic for two reasons. First, the chosen SNPs may not be causal for the outcome. This can happen, for instance, by using a noncausal SNP that is in high LD with a causal SNP. Second, because many traits are polygenic or omnigenic, multiple causal SNPs may be in high LD together. LD clumping is likely to omit one or more of the causal SNPs, which can diminish statistical power in MR analysis. In support of this is the finding that including multiple correlated SNPs in transcriptome-wide association studies improves power compared to using independent SNPs only [64–67]. Consequently, many MR methods have been recently developed to model multiple correlated instruments. 
[62] developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation. However, many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to be statistically independent. This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [61].

>>>>

## Methods with Summary data
The proliferation of publicly available GWAS data accelerates the development of MR methods with multiple SNP instruments that make use of GWAS summary statistics for model fitting. Our timeline highlights some of the many multi-instrument MR methods with GWAS summary statistics, with an increasing density of methods over time in the last three years (Figure 2). These methods include the inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted median estimation [@bowden_consistent_2016], Bayesian weighted MR [@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix [@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID [@yuan_likelihood_2021]. These methods differ in their approaches to three considerations that include instrument selection (from among the available SNPs), modeling and controlling for horizontal pleiotropy, and statistical inference procedure. We summarize some of these MR methods that use independent or correlated SNP instruments below. We will cover the remaining MR methods that model horizontal pleiotropy in a later section.

Figure 2: Two-sample MR with GWAS summary statistics methods publications density increases over time.
@burgess_mendelian_2013 termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from  independent SNPs by using Equation (5.1).

@bowden_consistent_2016, noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
weighted median-based estimators, building on research from @han2008detecting. Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [@hampel1971general]. Specifically, let  denote the  ordered ratio estimate statistic, from least to greatest. If an odd number, say , of SNP instruments is used, then the simple median estimator is defined as the  ordered ratio estimate. For an even number, say , number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, .
Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted median estimator. To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator. The weighted median estimator, then is the median of a distribution defined by having  as the  percentile, where  is the weight assigned to the  ordered ratio estimate. In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the penalized weights, , where  is the p-value resulting from the comparison of  to a  distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011].  Consequently, many MR methods have been recently developed to model multiple correlated instruments. 
@burgess_combining_2016 developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation. However, many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to be statistically independent. This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [@burgess_mendelian_2013].

## Assumption violations and Sensitivity Analysis
It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations. Let’s first consider the exclusion restriction assumption (assumption 2 above). It states that the SNP instrument must affect the outcome only through the exposure. The causal diagram in Figure  illustrates a scenario that violates this assumption [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, if the intermediate is measured and treated as the exposure, then an unbiased estimate of causal effect is possible, since the intermediate captures the two pathways, intermediate -> outcome and intermediate -> exposure -> outcome, through which the SNP instrument affects the outcome Y [@vanderweele_methodological_2014;@textor_graphical_2011;@textor_robust_2016].

Figure 3: Causal diagrams for MR. A. Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure. Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption. B. Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently.

Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that enable assumption relaxations and has motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight into how the results might differ if the assumptions be violated. Because not all confounding variables are known or measured, assumptions 2 and 3 are not fully verifiable in MR studies. Recognition of this fact has inspired the development of sensitivity analysis tools. While we can’t assess whether a SNP instrument is associated with every confounder of the exposure-outcome association, it is possible to examine the associations of the SNP instrument with the measured covariates. While absence of such associations doesn’t guarantee satisfaction of the assumption, presence of SNP-covariate associations must be investigated carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].
@burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when working with GWAS summary statistics. @burgess_sensitivity_2017 presented methods for both assessing the MR assumptions, to the extent possible, and performing robust analyses. For example, they used measured covariates to assess for possible associations with the SNP instruments. While they can’t rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.
To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of C-reactive protein (CRP) levels on coronary artery disease risk with four genetic variants in the CRP gene region and 17 other genetic variants that affect coronary artery disease risk. They used measured covariates to probe for SNP-covariate associations, and followed it up with scatter plots and Cochran’s Q test on the causal estimates to inquire about whether the SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003]. Additionally, @burgess_sensitivity_2017 suggested using a funnel plot, like those in the meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011]. Additionally, Egger regression can be useful in this setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].
@spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant pleiotropy assumption.
MR-GENIUS is a framework for two-sample MR analysis when individual-level data are available [@tchetgen_tchetgen_genius_2021]. @tchetgen_tchetgen_genius_2021 focused efforts on relaxing the exclusion restriction assumption and building on existing G-estimation methods to create an estimator that is robust to violations of the exclusion restriction and to additive unmeasured confounding [@robins1994correcting]. They observe that their estimator, in some settings, reduces to that of @lewbel2012using, which is widely used in econometrics, but not in the MR literature.

## Explicit modeling of horizontal pleiotropy
Besides sensitivity analysis, several methods have been developed to directly model horizontal pleiotropy to ensure the validity of MR assumptions. Pleiotropy, where a single genetic variant affects multiple traits, has a long history of study in genetics and complex traits [@solovieff_pleiotropy_2013;@visscher_plethora_2016;@stearns_one_2010]. In MR setting, horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that bypasses the exposure variable [@verbanck_detection_2018]. The presence of horizontal pleiotropy constitutes a violation of the standard MR assumptions and can lead to biased causal effect estimates and diminished statistical power. @verbanck_detection_2018 documented widespread horizontal pleiotropy in GWAS. We discuss approaches to modeling and accounting for horizontal pleiotropy below.
In the short time since publication of @verbanck_detection_2018, researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the SNP-exposure relationships. The second type of horizontal pleiotropy manifests in the presence of unobserved exposure-outcome confounding. It induces correlation between horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy violate standard MR modeling assumptions and can bias causal effect estimates and can increase false discoveries [@yuan_likelihood_2021]. Early MR analyses avoided confounding from horizontal pleiotropy by discarding instrumental SNPs that might be associated with the outcome [@verbanck_detection_2018;@zhu_causal_2018]. More recent methods have attempted to model horizontal pleiotropy [@yuan_testing_2020;@morrison_mendelian_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly challenging because the MR model likelihood often involves an integral that can’t be solved analytically. Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
@bowden_mendelian_2015 adapted Egger regression, originally developed to detect bias in meta-analyses, to detect bias from horizontal pleiotropy in two-sample MR studies with GWAS summary statistics or individual-level data [@egger_bias_1997]. Their approach is termed MR-Egger. They formulate their model as

where  designates confounders,  and  denote the exposure and outcome, respectively,  is a SNP instrument, and  and  are the zero-mean, normally distributed random errors.
@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument  and the effects defined in Equation (5.2) and define the new parameter  (Equation (5.3)).

In Equation (5.3),  when SNP  is a valid instrument.
MR-Egger fits the regression model in Equation (5.4).

The notation in Equation (5.4) could be a little misleading. The independent and dependent variables are  and , respectively, while the regression parameters are  and .
The slope coefficient  estimates the causal effect of the exposure on the outcome. Under the assumption that the SNP-exposure relationship is independent of the SNP’s pleiotropic effects, Egger’s test offers a valid test of the null causal hypothesis and consistently estimates the causal effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].
In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity  test. In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate  in Equation (5.5).

In Equation (5.5),  is the inverse variance of the Wald estimator, and .
While the  test tends to be conservative in small sample sizes, its power increases with increasing sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].

@hemani_orienting_2017, working in the setting of two-sample MR with summary GWAS data, recognized that measurement error in outcome and exposure can give misleading results with the inferred causal arrow in the wrong direction. To work around this issue, they extended MR with a method that infers the causal direction between two traits. Specifically, @hemani_orienting_2017 adapt Steiger’s Z test for a difference in correlations [@steiger1980tests] to assess which of the two traits in an analysis has the stronger correlation with the SNP instrument. That with the stronger correlation is treated as the exposure. @hemani_orienting_2017 applied their test in a study of DNA methylation and gene expression, where either direction of causality is plausible, and found that many methylation traits cause changes in gene expression.
@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data. They first calculate Wald ratio estimates for every instrumental SNP. They then apply smoothing to the empirical distribution of ratio estimates. The mode of this smoothed distribution is the simple mode-based estimate of the causal effect [@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by weighting the empirical distribution of ratio estimates by the inverse variance of each estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based estimators with weighted and unweighted median-based estimators, MR Egger regression, and IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.
@rees_robust_2019 approached the challenge of horizontal pleiotropy by developing three new methods for two-sample MR analysis with GWAS summary data: robust regression, penalized weights, and LASSO penalization. The first two can be viewed as modifications of MR-Egger and IVW methods. Together, they offer three strategies for downweighting or excluding variants with heterogeneous causal estimates. Recall that MR-Egger offers consistent causal effect estimates when there is no horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the “Instrument Strength Independent of Direct Effect” (InSIDE) assumption [@bowden_mendelian_2015]. The InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to formulate an estimator with high breakdown point and high efficiency. The second estimation method, with penalized weights, is much like the penalized median estimator, from above, except that the second argument in the minimum function is , instead of the  above. Finally the LASSO-based method fashions an estimator by applying a penalty to the objective function from MR-Egger to get Equation (6.1) as the objective function.

@rees_robust_2019 studied two procedures for choosing : a heterogeneity stopping rule and a cross-validation rule. Taken together, these three methods offer additional contributions to a suite of sensitivity analysis tools that practitioners should consider.
@foley_clustering_2021 recognized that in some MR studies, the Wald ratio estimates for different SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms and may identify the SNPs that are involved in them. They devised a method to detect these clusters in the collection of Wald ratio estimates and found SNP instruments that point to distinct causal pathways of biological importance [@foley_clustering_2021].
@cheng_mr-ldp_2020 sought a probabilistic model that accounts for both linkage disequilibrium among SNP instruments and horizontal pleiotropy. They termed their method MR-LDP. Using an approach from @zhu_bayesian_2017, @cheng_mr-ldp_2020 derived an approximate likelihood for their regression models of the exposure and outcome on the SNP instruments. Incorporation of a random effect into their models accounted for the variance in causal effect estimates due to horizontal pleiotropy. They situate inference within an empirical Bayesian framework and present a parameter-expanded variational Bayes expectation-maximization algorithm for estimation [@liu_parameter_1998]. By jointly modeling the distribution of GWAS summary statistics and causal effects, @cheng_mr-ldp_2020 efficiently accommodates multiple SNP instruments in linkage disequilibrium. One limitation of MR-LDP
is its reliance on the InSIDE assumption and resulting inability to accommodate correlated horizontal pleiotropy.
@zhu_iterative_2021 present IMRP, a method for causal effect estimation and horizontal pleiotropy detection. They describe a test to distinguish vertical pleiotropy from horizontal pleiotropy. By combining this test with two-sample MR analysis of GWAS summary statistics, they develop an iterative procedure for estimating the causal effect of exposure on outcome and test for presence of horizontal pleiotropy in the SNP instruments. The iterative procedure alternates between 1) updating the estimate of the causal effect and 2) testing for horizontal pleiotropy with the included SNP instruments. At each test for horizontal pleiotropy, some SNP instruments may be discarded if they demonstrate evidence of horizontal pleiotropic effects. The procedure ends when there remain only SNP instruments that don’t exhibit horizontal pleiotropy. In the authors’ simulations and data analysis sections, their new method performs about as well as GSMR, IVW, and MR-PRESSO in the settings with and without balanced pleiotropy and with and without satisfaction of the InSIDE assumption [@zhu_iterative_2021]. Finally, the authors note that IMRP is three orders of magnitude faster than simulations-based MR-PRESSO.
@grant_efficient_2020 build on research from @lin2015regularization to present a method that applies a  penalty to coefficients of covariates in two-sample MR with GWAS summary statistics. Importantly, their method doesn’t penalize the SNP-exposure association. Instead, it shrinks towards zero those coefficients for covariates that demonstrate little pleiotropy. In so doing, they extend work from @burgess_multivariable_2015 and @burgess_re_2015 to get a causal effect estimator that needs no valid SNP instruments and is robust to measured vertical pleiotropy and more statistically efficient than previous MR methods.
@xu_robust_2021 consider two classes of horizontal pleiotropy with their method, MRCIP. They partition horizontal pleiotropic effects into correlated or idiosyncratic horizontal pleiotropy. Idiosyncratic horizontal pleiotropy is that with a large effect size, while correlated horizontal pleiotropy, like its definition elsewhere in our manuscript, refers to correlation between the effects of the SNP instruments on the exposure and the effects of the SNP instruments on the outcome. MRCIP relaxes both the exclusion restriction assumption and the InSIDE assumption by directly modeling the correlated horizontal pleiotropy with random effects. Furthermore, to accommodate the idiosyncratic pleiotropy, the authors downweight SNP instruments that demonstrate strong direct effects on the outcome. They find that their MRCIP provides valid causal inference even when there are no valid SNP instruments (i.e., when all SNP instruments are invalid) and when the InSIDE assumption is violated [@xu_robust_2021]. Their PRW-EM algorithm adds a reweighting step to the traditional expectation maximization algorithm to estimate the causal effect parameter in MR analysis [@dempster1977maximum]. Finally, they compare MRCIP to other MR methods with both simulated and real data sets. @xu_robust_2021, like the developers of MR-RAPS, assume that the horizontal pleiotropy is balanced, i.e., that the mean horizontal pleiotropic effect is zero. One direction for future study is to augment MRCIP to permit unbalanced horizontal pleiotropy.
@iong_latent_2020 build on other research using mixture models in the two-sample MR setting with GWAS summary statistics. Like MR-Clust, MR-PATH, as the method from @iong_latent_2020 is called, works with the Wald ratio estimates from multiple SNP instruments. Coining the term “mechanistic heterogeneity” to refer to the distribution of Wald ratio estimates due to SNP involvement in distinct causal pathways, @iong_latent_2020 specify a -component normal mixture model. They use an expectation maximization algorithm to fit the model parameters (the component weights and the parameters for each of the  normal distributions) and suggest a Bayesian information criterion-based approach to choosing  [@schwarz1978estimating].
@bucur_inferring_2020 developed BayesMR, a Bayesian framework for two-sample MR with individual-level data and discuss an approximation for use with
GWAS summary statistics. Unlike many MR methods, BayesMR both aims to control for horizontal pleiotropy and to quantitatively assess the possibility of reverse causation, where the putative exposure is actually caused by the putative outcome. For  independent SNP instruments, @bucur_inferring_2020 permit each SNP instrument to have both a direct effect on the exposure and a horizontally pleiotropic direct effect on the outcome. Instrument effects and horizontally pleiotropic effects are assumed independent. Note that this is the Bayesian analog of the InSIDE assumption. After specifying the model, they present a nested sampling scheme [@skilling2006nested;@handley2015polychord1;@handley2015polychord] that simultaneously computes the model evidence for both causal directions (i.e., exposure causes outcome and outcome causes exposure) and acquires samples from the posterior distribution for Bayesian statistical inferences.
@sanderson_examination_2019 found that multivariable MR, where multiple exposures jointly cause an outcome, consistently estimates the effect of the exposures on the outcome. They examined multivariable MR with both individual-level data and GWAS summary statistics. @sanderson_examination_2019 also develop a generalized version of Cochran’s Q test for quantifying SNP instrument strength and validity. However, this test requires knowledge of the covariances between the effects of the SNP instruments on the exposures. While these covariances could be estimated from individual-level data, such data are often unavailable in the GWAS setting. Importantly, the authors note that the estimands for multivariable MR nad MR differ; multivariable MR estimates the direct causal effects of each exposure on the outcome, while MR estimates the total causal effect of an exposure on the outcome.
    1. Use GCTA-COJO to obtain SNP instruments [@genetic_investigation_of_anthropometric_traits_giant_consortium_conditional_2012]

    2. Prune all SNPs in linkage disequilibrium with the SNP instruments to obtain a set of “tag” SNPs with correlations less than 0.95

    3. Solve for the causal effect estimate with ridge regression of the outcome on the matrix resulting from concatenation of the SNP instrument genotypes matrix and the tag SNPs genotypes matrix.
Note that in step 3, ridge regression is preferred to ordinary least squares due to the limited statistical power of ordinary least squares in this setting. Presumably, this is due to collinearity in the matrix of tag SNP genotypes.

## Correlated horizontal pleiotropy

The second type of horizontal pleiotropy, from above, is sometimes called “correlated horizontal pleiotropy.” Correlated horizontal pleiotropy occurs when a SNP affects both the exposure and a variable that confounds the relationship between exposure and outcome (Figure 3B).
CAUSE and MRAID are two methods that aim to model correlated horizontal pleiotropy [@morrison_mendelian_2020;@yuan_likelihood_2021]. Their approaches to modeling correlated horizontal pleiotropy differ in important ways that ultimately affect performance. CAUSE allows for a proportion of SNPs to exhibit correlated pleiotropy, which @morrison_mendelian_2020 models as an effect on a shared, unobserved factor. The remaining SNPs are independent of this unobserved factor. Every SNP can have a nonzero pleiotropic effect on the outcome, and these pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on adaptive shrinkage from @stephens_false_2016. Finally, @morrison_mendelian_2020 compares two models to determine whether the GWAS summary statistics are consistent with a causal effect of the
exposure on the outcome. Specifically, they estimate the difference in the expected log pointwise posterior density for the model with causal effect fixed at zero and the model that permits a nonzero causal effect [@vehtari_practical_2017].
@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021]. These observations motivated @yuan_likelihood_2021 to develop a new method, MRAID. MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations (7.1) and (7.2)).

The estimated marginal effects for the  SNP instruments on the exposure are denoted by , while  represents the estimated effects of the same SNPs on the outcome. The  by  SNP instrument correlation matrices for the exposure and outcome are written as  and , respectively.  and  can be estimated with 1000 Genomes Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry [@the_1000_genomes_project_consortium_1000_2012]. The -vector error terms,  and , follow multivariate normal distributions with mean zero and variances  and , respectively. @yuan_likelihood_2021 then constructs a Gibbs sampler to make likelihood-based inferences [@geman1984stochastic]. To facilitate computations, the investigators make several assumptions about the collection of SNPs:
    1. Relatively small proportion of SNPs have nonzero effects on the exposure

    2. A relatively small proportion of SNPs demonstrate horizontal pleiotropy

    3. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs

    4. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.
By encoding these assumptions in prior distributions, @yuan_likelihood_2021 enables the inferential procedures to accommodate the observed data in the context of the assumptions on SNPs.

## Practice recommendations

## Binary, Count, and Time-to-event outcomes and exposures

While we’ve focused above on continuous outcomes and continuous exposures, some data are more naturally treated as binary, count, or time-to-event variables. MR methods for non-continuous outcomes and exposures is an active area of research. @burgess_improving_2012 and @allman_novel_2021 developed MR strategies for binary exposures and binary outcomes. @burgess_improving_2012 extended methods from @wald_fitting_1940 by fitting two logistic regressions, one for the outcome and one for the exposure. They then calculated the causal effect estimator as the ratio of the two logistic regression coefficient estimators. @allman_novel_2021 devised a new MR method for binary outcome and binary exposure by drawing on @gauvreau_analysis_1997 and treating the exposure and outcome as correlated binary random variables. They implemented an iterative optimization algorithm to simultaneously infer the causal effect parameter and other parameters.
Researchers have also leveraged other generalized linear models for analysis in MR studies [@nelder1972generalized]. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and power, with simulations of count random variables and binary random variables. With growing interest in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].
Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017]. @guo_body_2017, for example, studied post-diagnosis survival time in women with breast cancer. They treated survival as a time-to-event outcome, which enabled them to model it with a proportional hazards model [@cox_regression_1972] in two-stage MR with individual-level data. Similarly, @he_causal_2018, in a study of cardiovascular disease, applied proportional hazards regression when examining time to stroke as an outcome variable. In very recent work, @tchetgen_tchetgen_genius_2021 reported methods, MR GENIUS, that are robust to violations of the exclusion restriction assumption (under many assumed MR data generating processes). They considered binary, continuous, and time-to-event outcomes within their framework. We are unaware of MR studies that treat the exposure variable as a time-to-event, but it’s conceivable that future research will consider time-to-event traits as exposures in MR analyses.
